---
title: "Getting Started with FESLtelemetry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with FESLtelemetry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## Overview

FESLtelemetry is the publication-level analytical layer of the PositionR acoustic telemetry ecosystem. It is designed to work downstream of [positionRtools](https://github.com/jakebrownscombe/positionRtools), which handles data ingestion, quality control, filtering, and mortality detection.

```text
positionRtools  →  Load / QC / Filter / Mortality
                              ↓  (atel object)
glatos_exploreR →  Interactive exploration  →  [optional] filter_bookmark.json
                              ↓
FESLtelemetry   →  Residency / Network analysis / Publication figures / Templates
```

The package provides:

- **Station residency analysis** — station-transition tracking with date-level granularity
- **Movement network analysis** — pairwise movement matrices and spatial network plots
- **Analysis templates** — standardized scripts that define the FESL lab workflow
- **atel compatibility** — all core functions accept positionRtools `atel` objects directly, with backwards compatibility for plain GLATOS-format dataframes

---

## Installation

```{r install, eval = FALSE}
# FESLtelemetry
remotes::install_github("FishEcologyScience/FESL_package_telemetry")

# positionRtools (recommended — required for atel workflows)
remotes::install_github("jakebrownscombe/positionRtools")
```

```{r load}
library(FESLtelemetry)
library(tidyverse)
```

---

## Example Data

FESLtelemetry includes a built-in dataset for demonstrations: 100,000 acoustic telemetry detections from Hamilton Harbour in GLATOS format, along with a shoreline shapefile.

```{r example-data}
# Load built-in data
data("example_data_raw_dets", package = "FESLtelemetry")
data("shapefile_HH_WGS84",    package = "FESLtelemetry")

# Inspect structure
dplyr::glimpse(example_data_raw_dets)
```

```{r example-data-summary}
cat("Detections:", format(nrow(example_data_raw_dets), big.mark = ","), "\n")
cat("Animals:   ", length(unique(example_data_raw_dets$animal_id)), "\n")
cat("Stations:  ", length(unique(example_data_raw_dets$station_no)), "\n")
cat("Date range:", format(min(example_data_raw_dets$date), "%Y-%m-%d"),
    "to", format(max(example_data_raw_dets$date), "%Y-%m-%d"), "\n")
```

---

## Working with atel Objects

The recommended workflow starts with a positionRtools `atel` object — a standardized container produced after QC and filtering. All FESLtelemetry core functions accept `atel` directly.

### Converting raw files to atel

If you are starting from raw GLATOS (or OTN, Fathom) data files, use `raw_to_atel()` to produce an `atel` object in one step. This requires positionRtools.

```{r raw-to-atel, eval = FALSE}
# From GLATOS RDS files (most common FESL use case)
atel_obj <- raw_to_atel(
  detections  = "path/to/glatos_detections.rds",
  deployments = "path/to/glatos_receivers.rds",
  animals     = "path/to/glatos_animals.rds",
  source      = "GLATOS"
)

# From OTN CSV/XLSX exports
atel_obj <- raw_to_atel(
  detections  = "detections.csv",
  deployments = "deployments.xlsx",
  animals     = "animals.xlsx",
  source      = "OTN"
)
```

### Loading a saved atel object

After positionRtools processing, `atel` objects are typically saved as RDS or Parquet files.

```{r load-atel, eval = FALSE}
# Load from RDS (standard)
atel_obj <- readRDS("path/to/processed_atel.rds")

# Load from Parquet (positionRtools format)
atel_obj <- positionRtools::read_atel("path/to/parquet_directory/")
```

---

## Station Residency

`calculate_residency()` tracks individual fish movements between receiver stations and computes the total time spent at each station per day. It identifies consecutive detections at the same station and sums their durations.

### With atel input (recommended)

Column mapping is handled automatically — `receiver_sn`, `detection_datetime_utc`, and `deploy_lon` are translated to FESL conventions internally.

```{r residency-atel, eval = FALSE}
df_residency <- calculate_residency(atel_obj, units = "hours")
```

### With plain GLATOS dataframe

```{r residency-glatos}
df_residency <- calculate_residency(
  data          = example_data_raw_dets,
  animal_col    = "animal_id",
  station_col   = "station_no",
  timestamp_col = "detection_timestamp_est",
  date_col      = "date",
  lat_col       = "deploy_lat",
  long_col      = "deploy_long",
  units         = "hours"
)

dplyr::glimpse(df_residency)
```

### Interpreting residency output

Each row is one fish at one station on one day. The `residence` column holds total time in the specified units.

```{r residency-summary}
# Per-fish summary
df_residency %>%
  dplyr::group_by(animal_id) %>%
  dplyr::summarize(
    total_hours  = round(sum(residence, na.rm = TRUE), 1),
    n_stations   = dplyr::n_distinct(station_no),
    n_days       = dplyr::n_distinct(date),
    .groups      = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(total_hours)) %>%
  head(10)
```

### Residency heatmap

A common publication figure shows residency as a spatiotemporal heatmap — stations ordered geographically on the y-axis, date on the x-axis.

```{r residency-heatmap, fig.height = 6}
# Station coordinate key
temp_station_key <- example_data_raw_dets %>%
  dplyr::select(station_no, deploy_lat, deploy_long) %>%
  dplyr::slice_head(by = station_no)

df_residency %>%
  dplyr::left_join(temp_station_key, by = "station_no") %>%
  dplyr::mutate(
    station_factor = forcats::fct_reorder(
      as.factor(station_no), deploy_long, .fun = median
    )
  ) %>%
  ggplot2::ggplot(aes(x = date, y = station_factor, fill = residence)) +
  ggplot2::geom_tile() +
  viridis::scale_fill_viridis_c(end = 0.85, na.value = "grey90") +
  ggplot2::labs(
    x    = "Date",
    y    = "Station (ordered west to east)",
    fill = "Daily residency\n(hours)"
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))

rm(temp_station_key)
```

---

## Movement Network Analysis

`network_summary()` builds a movement network from detection data: it identifies when each fish moves from one receiver to another (using run-length encoding to collapse repeat detections) and assembles a pairwise movement matrix with spatial coordinates.

### With atel input (recommended)

```{r network-atel, eval = FALSE}
network_data <- network_summary(atel_obj)
```

### With plain GLATOS dataframe

```{r network-glatos}
network_data <- network_summary(
  data       = example_data_raw_dets,
  FishID     = example_data_raw_dets$animal_id,
  ReceiverID = example_data_raw_dets$station_no,
  lat        = example_data_raw_dets$deploy_lat,
  long       = example_data_raw_dets$deploy_long
)
```

### Inspecting the network object

`network_summary()` returns a named list with four components.

```{r network-inspect}
# Receiver locations with detection frequencies
head(network_data$receiver.locations)

# Individual fish movements (from → to)
head(network_data$individual.moves)

# Total movements between each pair
network_data$moves.matrix[1:5, 1:5]
```

### Plotting the network

`network_plot()` renders the network spatially. Node size is fixed; node colour reflects detection frequency (viridis scale). Edge width scales with movement frequency. An `sf` shapefile can be overlaid for geographic context.

```{r network-plot}
network_plot(
  data        = network_data,
  Min.traffic = 2,
  shapefile   = shapefile_HH_WGS84,
  labels      = TRUE,
  label.size  = 2.5,
  plot.title  = "Hamilton Harbour — Movement Network"
)
```

The plot is printed to the active device and returned invisibly as a ggplot object, so it can be captured for `ggsave()`:

```{r network-save, eval = FALSE}
p <- network_plot(network_data, shapefile = shapefile_HH_WGS84)
ggplot2::ggsave("network_plot.png", p, width = 10, height = 8, dpi = 300)
```

---

## Analysis Templates

Templates are standardized R scripts that encode the FESL lab workflow. Deploy them to your project directory with `use_template()`, then open, customize, and run.

```{r templates-list}
list_templates()
```

### Template 00 — Load and filter

The entry-point template for every FESL analysis session. Loads an `atel` object and optionally applies a filter bookmark exported from glatos_exploreR.

```{r template00, eval = FALSE}
use_template("load_and_filter", output_dir = "02_scripts")
```

The deployed script provides three paths:

1. **Path A** — load from RDS (standard, used 9/10 times)
2. **Path B** — load from Parquet files
3. **Optional filter block** — reads a `filter_bookmark.json` and re-applies the UI-selected filters (species, date range, lake, project) using positionRtools filter functions

### Template 01 — Network and residency

Performs station residency calculations and movement network analysis on the `atel` object produced by template00. Outputs `df_residency`, `network_data`, and a `plots` list with figure-ready ggplot objects.

```{r template01, eval = FALSE}
use_template("network_residency", output_dir = "02_scripts")
```

### Typical session workflow

```{r workflow, eval = FALSE}
# 1. Deploy templates to your project (once per project)
use_template("load_and_filter")
use_template("network_residency")

# 2. Open and fill in template00 — set param_atel_path, run
# 3. Open and run template01 — produces df_residency, network_data, plots
# 4. Customize plots for publication
ggplot2::ggsave("figures/fig2_network.png",
                plots$network, width = 10, height = 8, dpi = 300)
```

---

## Filter Bookmark Workflow

When working on a large multi-species, multi-lake dataset in glatos_exploreR, a user may identify a specific subset of interest interactively (e.g., walleye detections in Lake Ontario during 2023). The filter state can be exported as a `filter_bookmark.json` directly from the app.

Template00's optional filter block reads this file and re-applies the same filters to the full `atel` object using positionRtools:

```{r filter-bookmark, eval = FALSE}
# In template00, uncomment the filter bookmark block:
param_bookmark_path <- "filter_bookmark.json"
temp_filters <- jsonlite::fromJSON(param_bookmark_path)

# Filter to bookmarked species
temp_target_animals <- dplyr::filter(
  atel_obj$animals,
  species_common %in% temp_filters$filters$species
)
atel_obj <- positionRtools::filter_custom(
  atel_obj,
  animal_id %in% temp_target_animals$animal_id
)

# Filter to bookmarked date range
atel_obj <- positionRtools::filter_custom(
  atel_obj,
  detection_datetime_utc >= as.POSIXct(temp_filters$filters$date_start, tz = "UTC"),
  detection_datetime_utc <= as.POSIXct(temp_filters$filters$date_end,   tz = "UTC")
)
```

The `filter_bookmark.json` structure produced by glatos_exploreR:

```json
{
  "exported_from": "glatos_exploreR",
  "export_datetime": "2026-05-28T14:30:00",
  "filters": {
    "species":    ["walleye"],
    "lakes":      ["Lake Ontario"],
    "projects":   ["HH_2023"],
    "date_start": "2023-04-01",
    "date_end":   "2023-10-31"
  }
}
```

---

## Session Cleanup

All FESL templates follow Paul's lab conventions for environment cleanup. Intermediary objects use the `temp_` prefix and are removed at section boundaries:

```{r cleanup-pattern, eval = FALSE}
rm(list = ls(pattern = "^temp_"))
cat("Cleanup complete.\n")
```

---

## Further Reading

- **positionRtools** — data ingestion, QC, filtering, mortality detection, and export: `vignette("01_getting_started", package = "positionRtools")`
- **glatos_exploreR** — interactive network-level exploration of GLATOS telemetry data
- FESL lab coding conventions: `wiki/articles/coding--r-conventions.md`
```
